---
title: "igraph Tutorial"
author: "Suparna"
date: "01/04/2020"
output: html_document
---
# A tutorial to learn the basics of "igraph" package in R

install.packages("igraph")
install.packages("igraphdata")

```{r}
library("igraph")
library("igraphdata")
```

## Directed Graph

Create a directed graph with nodes that connect successive pairs.
N1->N2, N2->N3, N3->N4, N4->N1, N1->5, N2->N5 with 2 isolates N6, N7
```{r}
gr1_dir = graph(c("N1","N2","N2","N3","N3","N4","N4","N1","N1","N5","N2","N5"), isolates=c("N6","N7"))
```

Add attributes to vertices by calling V() function.  Nodes 1 to 5 are proteins, 7,8 are small molecules
```{r}
V(gr1_dir)$type <- c(rep("proteins",5),rep("small molecues",2))
```

Add edge weights, if you want a **weighted graph**, using E() function
```{r}
E(gr1_dir)$weight <- c(1.5, 2.0, 4.0, 2.5, 1.9, 1.0, 1.0)
plot(gr1_dir)
# We can also tamper with plot parameters

plot(gr1_dir, vertex.label.cex=0.8,vertex.label.dist=2.6, edge.arrow.size=0.5)

plot (gr1_dir,
vertex.label.cex = 0.8,
vertex.label.family="Verdana",
edge.width=E(gr1_dir)$weight*1.0,
edge.color="red")
```

## Undirected Graph

Create an undirected graph for same data.

```{r}
gr1_undir = graph(c("N1","N2","N2","N3","N3","N4","N4","N1","N1","N5","N2","N5"), isolates=c("N7","N8"), directed=F)
graph(c("N1","N2","N2","N3","N3","N4","N4","N1","N1","N5","N2","N5"), isolates=c("N7","N8"))
V(gr1_dir)$type <- c(rep("proteins",5),rep("small molecues",2))

# Plot the graph
plot(gr1_undir)
plot(gr1_undir, vertex.label.cex=0.8,vertex.label.dist=2.6, edge.arrow.size=0.5)

# Print the igraph objects gr1_dir and gr1_undir
print(gr1_dir)
print(gr1_undir)
```
**NOTE :** In the above printout, "U"->undirected, D-->"Directed", "N"-->named graph (edges named), "W"-> weighted graph", "B"-> bipartite, where nodes have names.

Above graph is "UN", Undirected and Named.
The two numbers 7,6 refer to 7 nodes, 6 edges

```{r}
# Print the vertices (nodes) and edges 
print(V(gr1_undir))
print(E(gr1_undir))

print(V(gr1_dir))
print(E(gr1_dir))

# Print the sparse network matrix
print(gr1_dir[])
print(gr1_undir[])
```

We can also access rows and columns of this matrix, as well as the attributes
```{r}
# Rows, columns
print(gr1_dir[1,])

# Attributes
print(  V(gr1_dir)$name )
print(  V(gr1_dir)$type )
print( E(gr1_dir)$weight)
```

Adding new vertices (nodes) and edges to the graph:

```{r}
#Add 2 new nodes "N8", "N9" to gr1_undir
gr1_dir <- gr1_dir + vertices(c("N8","N9"), type=c("gene","gene" ) )
print(V(gr1_dir))
print(vertex_attr(gr1_dir))

#Adding new edges that covers newly added vertices
gr1_dir <- gr1_dir + edges(c("N1","N8","N3","N9","N9","N8","N7","N1","N8","N7","N9","N7"),weight=c(1.5,2.0,3.0,1.0,1.0,1.0))
print(E(gr1_dir))

# Plot the new network

# gr1_dir <- delete_graph_attr(gr1_dir,"gene")
l = layout_on_grid(gr1_dir)
plot (gr1_dir,
vertex.label.cex = 0.8,
vertex.label.family="Verdana",
layout=l,
edge.width=E(gr1_dir)$weight*1.0,
edge.color="lightsteelblue")
```

## Computing the properties of the network

### Degree Measures

1. Edge density = ratio between edges in the graph to all pssoble edges between nodes

```{r}
print(  edge_density(gr1_dir) )
```

2. Degree of a node

```{r}
# Directed
deg <- degree(gr1_dir)
print(deg)

# Undirected
deg <- degree(gr1_undir)
print(deg)
```

3. Histogram of node degree
```{r}
hist(deg, breaks=10, main="Histogram of node degree")
```

4. Degree distribution and cumulative degree distribution
```{r}
deg <- degree(gr1_dir)
deg.dist = degree_distribution(gr1_dir,cumulative=T)

plot( x=0:max(deg), y=1-deg.dist, pch=19, cex=0.8, col="coral", xlab="Degree", ylab="Cumulative Frequency")
```

## Centrality Measures

1. Degree and Degree Centrality

```{r}
print (degree(gr1_dir))
print(centr_degree(gr1_dir))
```

2. Closeness and Closeness centrality (not well defined for disconnected graphs)

```{r}
print( closeness(gr1_dir) )
print( centr_clo(gr1_dir) )
```

3. Eigenvector Centrality
```{r}
print( eigen_centrality(gr1_dir)$vector )
print( centr_eigen(gr1_dir)$centralization )
```

4. Betweenness centrality
``` {r}
print( betweenness(gr1_dir) )
print( centr_betw(gr1_dir)  )
print( edge_betweenness(gr1_dir) )
```


## Distance Measures

Get all the shortest paths in the graph:
```{r}
# Gives a matrix of distances between two nodes or two sets of nodes
print( distances(gr1_dir) )
print( distances(gr1_undir) )
```

Get the average shortest path
```{r}
print(mean_distance(gr1_dir) )
print(mean_distance(gr1_undir) )
```

Examine the output of **shortest_path**
```{r}
# Single target node
selectedPaths1 <- shortest_paths(gr1_dir,from=V(gr1_dir)[name=="N1"],to = V(gr1_dir)[name=="N4"], output="both")
print(selectedPaths1)


# Multiple target nodes
selectedPaths2 <- shortest_paths(gr1_dir, from=V(gr1_dir)[name=="N1"], to = V(gr1_dir)[type=="genes"], output="both")
print(selectedPaths2)
```

All shortest paths with weight
```{r}
all_paths1 <- all_shortest_paths(gr1_dir,V(gr1_dir)[name=="N1"],to=V(gr1_dir)[name=="N5"])
print(all_paths1)
```


# Neighbourhood Functions

```{r}
# Edges of specific single node:
print (  incident(gr1_dir, V(gr1_dir)$name=="N3") )

# Edges from a specific set of nodes
print( incident_edges(gr1_dir, V(gr1_dir)$type=="proteins") )

# Neighbouring nodes of a single node
print( neighbors(gr1_dir, V(gr1_dir)$name=="N4") )


# Neighbouring nodes of a set of nodes
print( adjacent_vertices(gr1_dir, V(gr1_dir)$type=="proteins") )
```

## Order of the neighbour
The nodes within a distance of one edge from our node of interest are called first order neighbours. The ego function allows us to extract nodes further away from our node, within a specific order.

``` {r}
# Identify nodes within 2 orders from the node A
print(  ego(gr1_dir, order=2, nodes=V(gr1_dir)$name=="proteins")  )

## We can also select all the edges between two specific set of vertices.
print( E(gr1_dir)[V(gr1_dir)[type=="proteins"] %--% V(gr1_dir)[type=="gene"]] )
```